Panarchy: ripples of a boundary concept

### 2020-12-28 paper/paper_panarchy.Rmd
submission + pre-print preparation

---
title: "Panarchy: ripples of a boundary concept"
author: |
    | \small Juan Rocha^1,2,3^, Linda Luvuno^4^, Jesse Rieb^5^, Erin Crockett^6^,  Katja Malmborg^1^, Michael Schoon^7^, Garry Peterson^1^
    | \small
    | \small ^1^Stockholm Resilience Centre, Stockholm University, Kräftriket 2B, 10691 Stockholm
    | \small ^2^Future Earth, The Swedish Royal Academy of Sciences, Lilla Frescativägen 4A, 104 05 Stockholm
    | \small ^3^South American Institute for Resilience and Sustainability Studies, Maldonado, Uruguay
    | \small ^4^.Center for Complex Systems in Transition, Stellenbosch University, South Africa
    | \small ^5^Department of Geography, McGill University, 805 rue Sherbrooke ouest, Montréal (Québec) H3A 0B9, Canada
    | \small ^6^Department of Natural Resource Sciences, McGill University, 21111 Lakeshore Road, Sainte-11 Anne-de-Bellevue, QC, H9X 3V9, Canada
    | \small ^7^School of Sustainability, Arizona State University, Tempe, Arizona, USA

abstract: |
  How do social-ecological systems change over time?  In 2002 Holling and colleagues proposed the concept of Panarchy, which presented social-ecological systems as an interacting set of adaptive cycles, each of which is produced by the dynamic tensions between novelty and efficiency at multiple scales. Initially introduced as a conceptual framework and set of metaphors, panarchy has gained the attention of scholars across many disciplines and its ideas continue to inspire further conceptual developments. Almost twenty years after this concept was introduced we review how it has been used, tested, extended and revised.  We do this by using  qualitative methods and machine learning. We find that the adaptive cycle is the feature of panarchy that has attracted the most attention. Challenges remain in empirically grounding the metaphor, but recent theoretical and empirical work offer some avenues for future research.

bibliography: panarchy.bib
# csl: nature.csl
output:
  pdf_document:
    dev: pdf
    keep_tex: no
    toc: no
    latex_engine: xelatex
  word_document:
    toc: no
citecolor: blue
documentclass: article
font: Arial
fontsize: 10pt

linkcolor: blue
urlcolor: blue
header-includes:
- \usepackage{dcolumn, rotating, longtable, lineno, float, fontspec, array, tabularx}
- \setmainfont{Arial}
- \linenumbers
- \renewcommand{\abstractname}{}
- \graphicspath{{figures/}}
- \usepackage[margin=2.5cm]{geometry}
---

@@

Almost two decades ago, the edited book “Panarchy: Understanding Transformations in Human and Natural Systems”  [@Gunderson:2002vk], presented a synthetic perspective of how a group of leading social-ecological researchers associated with the Resilience Alliance understood change in social-ecological systems.  The concept of Panarchy was a key focus on this influential book. In the late 1990s and early 2000s, the Resilience Alliance was a productive, innovative highly collaborative group of environmental scientists that was focused on bridging social and natural sciences as well as theory and practice. They were very focused on addressing social-ecological problems by combining insights from the social and natural sciences, as well as the arts and humanities [@Parker:2012eo].

@@

Panarchy was, and still is, a boundary object that has enabled collaborations and nurtured new scientific frameworks. The ideas put forward have been applied in field studies, mathematical models, participatory work, and scenario development. Panarchy has inspired resilience assessments and guided decision making. In this article we pay tribute to the book by studying how the concepts and metaphors proposed have been further developed in the academic literature. We also document criticism of the concepts and identify key research frontiers.

@@

Panarchy posits that systems are organized in nested hierarchies across space and time, where each level of the hierarchy is a subsystem that can be in a different phase of the adaptive cycle. These phases are exploitation, conservation, release, and reorganization, the latter characterized by events of creative destruction (Fig 1). They are inspired by mathematical models used in economics and ecology, but broadly describe patterns of growth, collapse, and recovery that are common to populations, ecological communities, markets, or political organizations. While each subsystem in the hierarchy can be at a different phase of the cycle, these subsystems can influence each other through cross-scale interactions called “revolt” and “remember.”

The three-dimensional space has corners with attractors that can cause the adaptive cycle to stagnate: poverty and rigidity traps [@Gunderson:2002vk]. Poverty traps are described in panarchy as maladaptive states where potential, connectivity and resilience are low. Poverty traps are a series of feedback mechanisms that reinforce impoverished states [@Bowles:2006hx], limiting the system’s capacity to innovate and increase potential. The opposite corner, where potential, connectivity and resilience are high, is another maladaptive space called a rigidity trap. In that corner there is little space for experimentation and innovation. Examples include very conservative and bureaucratic systems.

@@

\begin{figure*}[h]
\centering
\includegraphics[width = 3in, height = 3in]{fig1_panarchy}
\caption{\textbf{Panarchy} is an heuristic of nested adaptive cycles that serves to represent a variety of systems and environmental problems. Adaptive cycles at different scales of the hierarchy can be connected through remember and revolt cross-scale interactions. Reprinted with permission from Island Press}
\label{fig:fig1}
\end{figure*}

@@

To answer these questions we combined an automated literature review based on topic modeling [@Blei:2012dk; @Griffiths:2004ey] with human-coded document analysis [@Bryman:2008va]. All data and code to replicate the analysis are available at: [https://github.com/juanrocha/panarchy](https://github.com/juanrocha/panarchy) and a public online repository (add figshare link when finished).

@@

To prepare the data for topic modeling, we constructed a document term matrix with documents in rows and words in columns. Here our unit of analysis for the document are the abstracts retrieved, and the matrix contains the count of words per document. We removed stop words (e.g. "the", "a") and digits from the matrix, as well as a list of words that were over represented in our data  and are common in the scientific literature but are unrelated to the papers’ topics(e.g. "paper", "study", "aim").

@@

*Document analysis*: We complemented our unsupervised approach with document analysis [@Bryman:2008va] by coding an additional set of categorical variables for a sample of highly cited papers (N = 41). We annotated qualitative aspects such as use of the adaptive cycle, identification of its phases, whether the paper is conceptual, modeling or an empirical study. When empirical, we record the temporal and spatial scales of the case study. We also identified what aspects of panarchy were most used in the papers. For example whether there is emphasis on cross-scale interactions, or poverty and rigidity traps. We used text annotations to capture potential criticisms as well as the methods used. The code book is available in the supplementary material (archived in XX repository).

@@

_Panarchy_, the book published in 2002, has been cited over 7200 times according to Google Scholar. The scientific paper that introduced the book to the scientific community [@Holling:2001cu] has received 1715 citations in the Web of Science, and 1923 in Scopus. Roughly half of the citations have come from environmental (28%) and social sciences (22%). Computer science (2.1%)and arts and humanities (2.6%) have been the least represented.

A first exploration on the data suggest that content related to the adaptive cycle or general resilience has slightly decreased, while issues on development and management had gained attention (Fig 2). This simple exploration is not necessarily the best fit for our data, but serves to exemplify how topic models work. A topic is a set of words that are ranked according to a probability that they represent an underlying content of the document (Fig 1A). For example, the words "resilience", "systems", and "social" have a high probability to capture the content of topic 5. Similarly, there is a cluster of papers whose content is largely dominated by topic 1 on the adaptive cycle (gamma > 0.6), featuring it as one of the most active areas of research despite its relative decline in recent years (Fig 2).

Comparing algorithm performance revealed that Gibbs sampling was the best fit for the data. As a rule of thumb, an ideal method should maximize entropy and the log-likelihood estimation while minimizing perplexity [@Grun:2011tb]. Gibbs sampling maximised both entropy and likelihood with our data when compared to other alternatives. The second best was the variational-expectation maximisation algorithm (VEM) when $\alpha$ was not set constant. $\alpha$ is a hyper parameter that weights the evenness of topic distribution. A lower $\alpha$ than default values indicates that the documents can be described by rather fewer topics, or that its distribution is very uneven. In fact, we observed that increasing the number of topics from 5 to 100 topics did increase entropy at expense of reducing $\alpha$, meaning that despite the larger number of topics, the main content of a document is still captured by a few of them. The log-likelihood maximization stopped at around 25-50 topics. Thus, we restricted the rest of our analysis to 25 topics.

\begin{figure*}[h]
\centering
\includegraphics[width = 4in, height = 3.5in]{fig2_topicnumber}
\caption{\textbf{Algorithm and number of topic selection} Gibbs sampling maximizes entropy and log-likelihood estimation, making it a suitable algorithm for our data (A). Increasing the number of topics (from 5-100) shows that $\alpha$ decreases, suggesting that despite the larger number of topics, a few of them suffice to describe most papers (B). Log-likelihood is maximized for 50 topics followed closely with 25. While 50 topics is marginally better, for visualization purposes we restrict our analysis to 25. Note that perplexity cannot be calculated for Gibbs sampling, hence the missing value in A and absence in B.}
\label{fig:algorithm}
\end{figure*}

A topic is a set of words that are ranked according to a probability that they represent an underlying content of the document (Fig 3). For example, the words "resilience", "vulnerability", “risk” and "disaster" have a high probability to capture the content of topic 25. Papers early in our time series (2001-3) have high content largely dominated by topics 25 on resilience, 24 on the adaptive cycle, and 18 on adaptive governance and management. Towards the end of the time series, topic 3 on urban systems and 9 on local communities and knowledge have spikes up to 9% of the content of each year (2019-2020). For comparison, if all topics were equally represented in the content, they would have 4% in the corpus (grey line in Fig 3B).

Despite fluctuations, most topics showed a relatively constant level of interest over time (Fig 3, Fig S2). We did not observe strong trends, but some topics have gained a small amount of attention. For example, topic 1 on sustainability, or topic 10 on ecosystem services appeared consistently across time. In contrast, research on business innovation (topic 20), urban infrastructure (topic 3), or archaeology (topic 5) have gained attention in recent years. Topic 8 was an outlier, with a selection of papers ranking high on content related to network infrastructure and performance, possibly from engineering disciplines. It was the only topic with a set of papers that was clearly distinct from the rest of the collection, and showed a decreasing trend over time.

\begin{figure*}[h]
\centering
\includegraphics[width = 5in, height = 5in]{fig3_topics25}
\caption{\textbf{Panarchy topics over time} The number of papers per year (A) with a maximum of 324 in 2019. The proportion of papers per year (B) and the relative proportion of topic content per year (C) do not show strong trends for the time window with most papers (gray area in A). Each topic is summarized in (D) by the top 10-words that best describe them according to the posterior probability of our  model fit. See FigS2 for a panel of time series per topics}
\label{fig:topics25}
\end{figure*}

The human-coded document analysis revealed that the most common feature of panarchy in the literature was the adaptive cycle (81%, N = 42) followed by cross-scale interactions (Fig 4). Poverty and rigidity traps were less studied features in our sample, even when accounting for slightly different terminology such as "lock-ins". The bulk of the papers analyzed were conceptual papers, many without a method or discernible research question. Roughly half of the papers were conceptual or only used panarchy as a metaphor. Over 40% of papers in our sample went a step further and used panarchy as an empirical construct, for example by attempting to identify the phases of the adaptive cycle (61%). Out of the 22 empirical cases, 6 were at the time scale of centuries, 7 on decades, 6 on years, and 1 in weeks. Spatially, 4 were at a city scale, 17 regional, and 1 national. Empirical papers were dominated by qualitative methods (77%, N = 22) and were generally retrospective historical reviews.

@@

\begin{figure*}[h]
\centering
\includegraphics[width = 6in, height = 3in]{fig4_qual_results}
\caption{\textbf{Qualitative results} Document analysis was used to disentangle the different panarchy features addressed by a sub sample of papers (N = 42). Over half of the papers are conceptual work, and most empirical papers fall into qualitative methods.}
\label{fig:qual}
\end{figure*}

# Discussion
(1000w)

@@

This paper aimed to study the use of panarchy related concepts and their evolution since the publication of the book. Systematic literature reviews often suffer from the limitation of restricting sample size to a subset of what is readable on the time frame of a project -- between dozens and maybe hundreds of papers. Qualitative analysis offers rich insights into the papers but is limited by sample size. In our analysis, the qualitative insights could be biased towards a non-random selection of papers that aligned our research interests, or highly cited, review type of work. Topic modeling enabled us to complement the analysis to all papers reported in major scientific databases. It has the advantage of reproducibility, and reduces bias, but offers little insights about the dimensions of panarchy used, methods, or criticisms. Here we show that both methods combined are a powerful combination. Future studies could benefit from including gray literature such as theses, non-governmental organizations reports, government agencies reports, or non-english literature. Previous studies have shown that the use of full text instead of abstracts can also offer additional insights on the automated analyses, for example in attributing impacts of ecosystem services from regime shifts in social-ecological systems [@Rocha:2015tk]. Replication attempts do however face the challenge of accessing full text when many of the papers are behind pay-walls.

# Conclusion
(400w)

We reviewed the academic literature of the last two decades to trace how Panarchy related concepts have evolved, in what type of problems they have been applied and found useful, and what remains as the key frontiers of research. Despite a growing body of literature, no topic seems to dominate the academic production of scholars using these concepts. The feature most used is the adaptive cycle, and the problems where it is found most useful is in studies with a historical perspective, either in short time horizons such as natural resource management, urban development, or conservation; all the way to archaeology and anthropology studies on the time horizon of centuries to millennia. Hierarchies in scale or maladaptive traps have received comparatively less attention. The frontiers of research point out to the problem of measurement and prediction. For example, how to observe and approximate resilience, connectedness and potential in the present; or how likely a current social-ecological system is to undergo a transition in the phases of the adaptive cycle, or a “memory” or “revolt” type of cross-scale interaction. While our survey of the literature offers a few options to empirically ground panarchy concepts, we believe the problem of measurement and prediction are fertile ground for future research efforts.

>>>>

---
title: "Panarchy: ripples of a boundary concept"
author: |
    | \small Juan Rocha^1,2,3^, Linda Luvuno^4^, Jesse Rieb^5^, Erin Crockett^6^,  Katja Malmborg^1^, Michael Schoon^7^, Garry Peterson^1^
    | \small
    | \footnotesize ^1^Stockholm Resilience Centre, Stockholm University, Kräftriket 2B, 10691 Stockholm
    | \footnotesize ^2^Future Earth, The Swedish Royal Academy of Sciences, Lilla Frescativägen 4A, 104 05 Stockholm
    | \footnotesize ^3^South American Institute for Resilience and Sustainability Studies, Maldonado, Uruguay
    | \footnotesize ^4^Center for Complex Systems in Transition, Stellenbosch University, South Africa
    | \footnotesize ^5^Department of Geography, McGill University, 805 rue Sherbrooke ouest, Montréal (Québec) H3A 0B9, Canada
    | \footnotesize ^6^Department of Natural Resource Sciences, McGill University, 21111 Lakeshore Road, 
    | \footnotesize Sainte-11 Anne-de-Bellevue, QC, H9X 3V9, Canada
    | \footnotesize ^7^School of Sustainability, Arizona State University, Tempe, Arizona, USA
abstract: |
  How do social-ecological systems change over time? In 2002 Holling and colleagues proposed the concept of Panarchy, which presented social-ecological systems as an interacting set of adaptive cycles, each of which is produced by the dynamic tensions between novelty and efficiency at multiple scales. Initially introduced as a conceptual framework and set of metaphors, panarchy has gained the attention of scholars across many disciplines and its ideas continue to inspire further conceptual developments. Almost twenty years after this concept was introduced we review how it has been used, tested, extended and revised.  We do this by combining qualitative methods and machine learning. Document analysis was used to code panarchy features that are commonly used in the scientific literature (N = 42), a qualitative analysis that was complemented with topic modeling of 2177 documents. We find that the adaptive cycle is the feature of panarchy that has attracted the most attention. Challenges remain in empirically grounding the metaphor, but recent theoretical and empirical work offer some avenues for future research.
bibliography: panarchy.bib
# csl: nature.csl
output:
  pdf_document:
    dev: pdf
    keep_tex: yes
    toc: no
    latex_engine: xelatex
  rtf_document:
    toc: no
citecolor: blue
documentclass: article
font: Arial
fontsize: 10pt
linkcolor: blue
urlcolor: blue
header-includes:
- \usepackage{dcolumn, rotating, longtable, lineno, float, fontspec, array, tabularx}
- \setmainfont{Arial}
- \renewcommand{\abstractname}{}
- \graphicspath{{figures/}}
- \usepackage[margin=2.5cm]{geometry}

## For Arxiv submission, use pdflatex as engine, comment out font, increase font size to 11, delete the package fontspec and the graphics path. Compile in latex so the bibliography is part of the tex file produced. 
---

@@

Almost two decades ago, the edited book “Panarchy: Understanding Transformations in Human and Natural Systems” [@Gunderson:2002vk], presented a synthetic perspective of how a group of leading social-ecological researchers associated with the Resilience Alliance understood change in social-ecological systems. The concept of Panarchy was a key focus on this influential book. In the late 1990s and early 2000s, the Resilience Alliance was a productive, innovative highly collaborative group of environmental scientists that was focussed on bridging social and natural sciences as well as theory and practice. They were very focussed on addressing social-ecological problems by combining insights from the social and natural sciences, as well as the arts and humanities [@Parker:2012eo].

@@

Panarchy remains a boundary object that has inspired research topics, enabled collaborations and nurtured new scientific frameworks. The ideas put forward have been applied in field studies, archaeology, mathematical models, participatory work, and scenario development. Panarchy has inspired resilience assessments and guided decision making. In this article we pay tribute to the book by studying how the concepts and metaphors proposed have been further developed in the academic literature. We also document criticism of the concepts and identify key research frontiers.

Panarchy proposes that it is useful to conceptualize systems in terms of interacting adaptive cycles. The adaptive cycle was an idea first proposed by Holling based upon his experience working and studying managed ecosystems [@biosphere:hJ4c_Ua8]. It was meant to be a conceptual tool that focussed attention on processes of destruction and reorganization, which had been neglected in comparison to those of growth and conservation. Considering these processes provides a more complex may to understand system dynamics.

An adaptive cycle alternates between long periods of system aggregation, connection, and accumulation, and shorter periods of disruption and reorganization. The adaptive cycle exhibits two major phases. The first, often referred to as the front-loop, from r to K, is the slow, incremental phase of growth and accumulation. The second, referred to as the back-loop, from Omega to Alpha, is the rapid phase of reorganization leading to renewal. A system going into the back-loop can either remerge in a similar form, or may re-emerge as a new type of system with new boundaries and key components. Resilience researchers proposed that tensions between demands for organization and efficiency versus demands for novelty and diversity drive adaptive cycle type dynamics in many different types of complex systems 
[@Holling:2002wu].

@@

\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 2.5in]{fig1_panarchy}
\caption{\textbf{Panarchy} is an heuristic of nested adaptive cycles that serves to represent a variety of systems and environmental problems. Adaptive cycles (A) at different scales of the hierarchy (B) can be connected through remember and revolt cross-scale interactions.}
\label{fig:fig1}
\end{figure*}

Panarchy posits that systems are organized in nested hierarchies across space and time, where each level of the hierarchy is a subsystem that can be in a different phase of the adaptive cycle. These phases are exploitation, conservation, release, and reorganization, the latter characterized by events of creative destruction (Fig. \ref{fig:fig1}). They are inspired by mathematical models used in economics and ecology, but broadly describe patterns of growth, collapse, and recovery that are common to populations, ecological communities, markets, or political organizations. While each subsystem in the hierarchy can be at a different phase of the cycle, these subsystems can influence each other through cross-scale interactions called “revolt” and “remember.”

The three-dimensional space has corners with attractors that can cause the adaptive cycle to stagnate: poverty and rigidity traps [@Gunderson:2002vk]. Poverty traps are described in panarchy as maladaptive states where potential, connectivity and resilience are low. Poverty traps are a series of feedback mechanisms that reinforce impoverished states [@Bowles:2006hx], limiting the system’s capacity to innovate and increase potential. The opposite corner, where potential, connectivity and resilience are high, is another maladaptive space called a rigidity trap. In that corner there is little space for experimentation and innovation. Examples include systems where ecological resilience has been expensively replaced by artificial processes to maintain the system such as levees, flood barriers, or chemical control of pests [@Holling:1996wx].

@@

@@

To answer these questions we combined an automated literature review based on topic modeling [@Blei:2012dk; @Griffiths:2004ey] with human-coded document analysis [@Bryman:2008va]. All data and code to replicate the analysis are available at: [https://github.com/juanrocha/panarchy](https://github.com/juanrocha/panarchy) and a public online repository ([10.6084/m9.figshare.13490919](10.6084/m9.figshare.13490919)).

@@

To prepare the data for topic modeling, we constructed a document term matrix with documents in rows and words in columns. Here our unit of analysis for the document are the abstracts retrieved, and the matrix contains the count of words per document. We removed stop words (e.g. "the", "a") and digits from the matrix, as well as a list of words that were over represented in our data  and are common in the scientific literature but are unrelated to the papers’ topics (e.g. "paper", "study", "aim").

@@

*Document analysis*: We complemented our unsupervised approach with document analysis [@Bryman:2008va] by coding an additional set of categorical variables for a sample of highly cited papers (N = 41). We annotated qualitative aspects such as use of the adaptive cycle, identification of its phases, whether the paper is conceptual, modeling or an empirical study. When empirical, we record the temporal and spatial scales of the case study. We also identified what aspects of panarchy were most used in the papers. For example whether there is emphasis on cross-scale interactions, or poverty and rigidity traps. We used text annotations to capture potential criticisms as well as the methods used. The code book is available in the supplementary material (archived in [10.6084/m9.figshare.13490919](10.6084/m9.figshare.13490919)).

@@

Panarchy, the book published in 2002, has been cited over 7200 times according to Google Scholar. The scientific paper that introduced the book to the scientific community [@Holling:2001cu] has received 1715 citations in the Web of Science, and 1923 in Scopus. Roughly half of the citations have come from environmental (28%) and social sciences (22%). Computer science (2.1%) and arts and humanities (2.6%) have been the least represented.

\begin{figure*}[ht]
\centering
\includegraphics[width = 5in, height = 4in]{fig2_topicnumber}
\caption{\textbf{Algorithm and number of topic selection} Gibbs sampling maximizes entropy and log-likelihood estimation, making it a suitable algorithm for our data (A). Increasing the number of topics (from 5-100) shows that $\alpha$ decreases, suggesting that despite the larger number of topics, a few of them suffice to describe most papers (B). Log-likelihood is maximized for 50 topics followed closely with 25. While 50 topics is marginally better, for visualization purposes we restrict our analysis to 25. Note that perplexity cannot be calculated for Gibbs sampling, hence the missing value in A and absence in B.}
\label{fig:algorithm}
\end{figure*}

Comparing algorithm performance revealed that Gibbs sampling was the best fit for the data (Fig \ref{fig:algorithm}). As a rule of thumb, an ideal method should maximize entropy and the log-likelihood estimation while minimizing perplexity [@Grun:2011tb]. Gibbs sampling maximised both entropy and likelihood with our data when compared to other alternatives. The second best was the variational-expectation maximisation algorithm (VEM) when $\alpha$ was not set constant. $\alpha$ is a hyper parameter that weights the evenness of topic distribution. A lower $\alpha$ than default values indicates that the documents can be described by rather fewer topics, or that its distribution is very uneven. In fact, we observed that increasing the number of topics from 5 to 100 topics did increase entropy at expense of reducing $\alpha$, meaning that despite the larger number of topics, the main content of a document is still captured by a few of them. The log-likelihood maximization stopped at around 25-50 topics. Thus, we restricted the rest of our analysis to 25 topics.

\begin{figure*}[ht]
\centering
\includegraphics[width = 5in, height = 5in]{fig3_topics25}
\caption{\textbf{Panarchy topics over time} The number of papers per year (A) with a maximum of 324 in 2019. The proportion of papers per year (B) and the relative proportion of topic content per year (C) do not show strong trends for the time window with most papers (gray area in A). Each topic is summarized in (D) by the top 10-words that best describe them according to the posterior probability of our model fit. See FigS1 for a panel of time series per topics}
\label{fig:topics25}
\end{figure*}

A topic is a set of words that are ranked according to a probability that they represent an underlying content of the document (Fig \ref{fig:topics25}). For example, the words "resilience", "vulnerability", “risk” and "disaster" have a high probability to capture the content of topic 25. Papers early in our time series (2001-3) have high content largely dominated by topics 25 on resilience, 24 on the adaptive cycle, and 18 on adaptive governance and management. Towards the end of the time series, topic 3 on urban systems and 9 on local communities and knowledge have spikes up to 9% of the content of each year (2019-2020). For comparison, if all topics were equally represented in the content, they would have 4% in the corpus (grey line in Fig \ref{fig:topics25}B).

Despite fluctuations, most topics showed a relatively constant level of interest over time (Fig \ref{fig:topics25}, Fig \ref{fig:figS1}). We did not observe strong trends, but some topics have gained a small amount of attention. For example, topic 1 on sustainability, or topic 10 on ecosystem services appeared consistently across time. In contrast, research on business innovation (topic 20), urban infrastructure (topic 3), or archaeology (topic 5) have gained attention in recent years. Topic 8 was an outlier, with a selection of papers ranking high on content related to network infrastructure and performance, possibly from engineering disciplines. It was the only topic with a set of papers that was clearly distinct from the rest of the collection, and showed a decreasing trend over time.

The human-coded document analysis revealed that the most common feature of panarchy in the literature was the adaptive cycle (81%, N = 42) followed by cross-scale interactions (Fig \ref{fig:qual}). Poverty and rigidity traps were less studied features in our sample, even when accounting for slightly different terminology such as "lock-ins". The bulk of the papers analyzed were conceptual papers, many without a method or discernible research question. Roughly half of the papers were conceptual or only used panarchy as a metaphor. Over 40% of papers in our sample went a step further and used panarchy as an empirical construct, for example by attempting to identify the phases of the adaptive cycle (61%). Out of the 22 empirical cases, 6 were at the time scale of centuries, 7 on decades, 6 on years, and 1 in weeks. Spatially, 4 were at a city scale, 17 regional, and 1 national. Empirical papers were dominated by qualitative methods (77%, N = 22) and were generally retrospective historical reviews.

@@

\begin{figure*}[ht]
\centering
\includegraphics[width = 5in, height = 2.5in]{fig4_qual_results}
\caption{\textbf{Qualitative results} Document analysis was used to disentangle the different panarchy features addressed by a sub-sample of papers (N = 42). Over half of the papers are conceptual work, and most empirical papers fall into qualitative methods.}
\label{fig:qual}
\end{figure*}

# Discussion

@@

This paper aimed to study the use of panarchy related concepts and their evolution since the publication of the book. Systematic literature reviews often suffer from the limitation of restricting sample size to a subset of what is readable on the time frame of a project -- between dozens and maybe hundreds of papers. Qualitative analysis offers rich insights into the papers but is limited by sample size. In our analysis, the qualitative insights could be biased towards a non-random selection of papers that aligned our research interests, or highly cited, review type of work. Topic modeling enabled us to complement the analysis to all papers reported in major scientific databases. It has the advantage of reproducibility and reduces sample bias, but offers little insights about the dimensions of panarchy used, methods, or criticisms. Here we show that both methods combined are a powerful combination. Future studies could benefit from including gray literature such as theses, books, non-governmental organizations reports, government agencies reports, or non-english literature. Previous studies have shown that the use of full text instead of abstracts can also offer additional insights on the automated analyses, for example in attributing impacts of ecosystem services from regime shifts in social-ecological systems [@Rocha:2015tk]. Replication attempts do however face the challenge of accessing full text when many of the papers are behind pay-walls.

# Conclusion

We reviewed the academic literature of the last two decades to trace how Panarchy related concepts have evolved, in what type of problems they have been applied and found useful, and what remains as the key frontiers of research. Despite a growing body of literature, no topic seems to dominate the academic production of scholars using these concepts. The feature most used is the adaptive cycle, and the problems where it is found most useful is in studies with a historical perspective, either in short time horizons such as natural resource management, urban development, or conservation; all the way to archeology and anthropology studies on the time horizon of centuries to millennia. Hierarchies in scale or maladaptive traps have received comparatively less attention. The frontiers of research point out to the problem of measurement and prediction. For example, how to observe and approximate resilience, connectedness and potential in the present; or how likely a current social-ecological system is to undergo a transition in the phases of the adaptive cycle, or a “memory” or “revolt” type of cross-scale interaction. While our survey of the literature offers a few options to empirically ground panarchy concepts, we believe the problem of measurement and prediction are fertile ground for future research efforts.

# Acknowledgements

JCR acknowledges financial support from Formas research grant (942-2015-731), the Stockholm Resilience Centre, and Stockholm University.
